NumPy---Array Programming in Python

### 2020-01-26 body.tex
Rearrange ecosystem section

While NumPy provides a solid foundation for array programming, the scientific
Python ecosystem is composed of many different libraries, which build on and
extend this functionality.
These libraries can roughly be organized into three layers:
core libraries, technique-specific libraries, and domain-specific libraries.

The core libraries include SciPy, Matplotlib, Pandas, and xarray.
SciPy adds various mathematical, scientific, and engineering routines;
Matplotlib generates publication-ready figures and visualizations,
Pandas provides indexed arrays, known as dataframes; and
xarray implements arrays with labeled axes.

Employing these low-level array-aware functions are widely used
technique-specific libraries, including scikit-learn (machine learning),
scikit-image (image processing), statsmodels (statistics),  networkx (graphs
and complex networks) \cite{SciPyProceedings_11}, and Napari (volumetric image
visualization).

Finally, domain-specific libraries and applications are built on this diverse
foundation of tooling; these include Astropy (astronomy) \cite{astropy:2013,
astropy:2018}, Biopython (computational biology and bioinformatics) \cite{cock2009biopython},
NIPY (neuroimaging) \cite{millman2007analysis}, and Pangeo (earth science) \cite{2018EGUGA..2012146H}.

@@

The \tt{eht-imaging} library developed by the Event Horizon Telescope 
collabortion relies on many components of the scientific Python ecosystem.
NumPy arrays were used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
Scipy was used to provide tools used in general image processing such as 
filtering and image alignment, while scikit\_image provided higher-level
functionality such as edge filters and Hough transforms.
Other components of Scipy are used throughout the library, such as the
\tt{scipy.optimize} package for handling general optimization tasks.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.
% N.B. - most of this comes from introspecting the Image class defined in
% ehtim/image.py.

Tools for analyzing data from gravitational wave observatories such as LIGO
and Virgo are provided by the \tt{pycbc} package, which was used in the first
detection of gravitationaly waves \cite{abbott2016observation} and in 
on-going analysis of data from LIGO and Virgo.
\tt{pycbc} make extensive use of the scientific Python ecosystem.
The time-series data from the interferometers are stored in NumPy arrays,
while the \tt{scipy.signal} is used to construct and apply filters to the
data.
Matplotlib is used to visualize data at all points along the analysis chain
including raw data from the instrumentation and the time-frequency 
visualization of the ``chirp'' from the binary black hole merger.

Exposing array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.
The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.

>>>>

While NumPy provides a solid foundation for array programming, the scientific
Python ecosystem is composed of many different libraries, which build on and
extend this functionality \fixme{(Fig.)}.
SciPy provides fundamental algorithms for scientific computing on arrays
including many mathematical, scientific, and engineering routines.
SciPy and NumPy are closely related and were developed together often
by the same people.
Matplotlib generates publication-ready figures and visualizations.
Employing these core array-aware libraries are widely used
technique-specific libraries\cite{SciPyProceedings_11}.
Finally, several domain-specific
libraries \cite{astropy:2013,astropy:2018,cock2009biopython,millman2007analysis,2018EGUGA..2012146H}
are built on this diverse foundation of tooling.

Exposing array programming primitives, as well as the surrounding ecosystem of
tools, in Python---an interpreted language---creates an ideal environment for
interactive, exploratory data analysis where users may iteratively inspect,
manipulate, and visualize their data.
This has made the scientific Python ecosystem popular for scientific
research.

%These libraries can roughly be organized into three layers:
%core libraries, technique-specific libraries, and domain-specific libraries.

%SciPy adds various mathematical, scientific, and engineering routines;
%Matplotlib generates publication-ready figures and visualizations,
%Pandas provides indexed arrays, known as dataframes; and
%xarray implements arrays with labeled axes.

@@

For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration relies on many components of the scientific Python ecosystem.
NumPy arrays were used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image 
reconstruction.
Scipy was used to provide tools used in general image processing such as 
filtering and image alignment, while scikit\_image provided higher-level
functionality such as edge filters and Hough transforms.
Other components of Scipy are used throughout the library, such as the
\code{scipy.optimize} package for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11} was used for image comparisons.
Astropy \cite{astropy:2013, astropy:2018} is used in quite a few places,
primarily for I/O and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including the generation of the final image of the black hole.
% N.B. - most of this comes from introspecting the Image class defined in
% ehtim/image.py.

Tools for analyzing data from gravitational wave observatories such as LIGO
and Virgo are provided by the \code{pycbc} package, which was used in the first
detection of gravitationaly waves \cite{abbott2016observation} and in 
on-going analysis of data from LIGO and Virgo.
\code{pycbc} make extensive use of the scientific Python ecosystem.
The time-series data from the interferometers are stored in NumPy arrays,
while the \code{scipy.signal} is used to construct and apply filters to the
data.
Matplotlib is used to visualize data at all points along the analysis chain
including raw data from the instrumentation and the time-frequency 
visualization of the ``chirp'' from the binary black hole merger.

The rich ecosystem of tools building on NumPy and the larger community
of scientists and programmers using those tools mean that NumPy has a
responsibility to maintain considerable stability.
This need creates interesting challenges for NumPy development as the
computational environment scientists are analyzing their data on continues
to evolve and change.
